---
title: "Econometrics: R introduction"
description: ""
author:
  - name: Basil Okola
    url: https://bokola.github.io/
    # orcid: 0000-0002-5300-3075
    affiliation: Master of Statistics and Data Science @ Hasselt University
    affiliation-url: https://ucsb-meds.github.io/ 
date: 2026-01-21
categories: [econometrics, instrumental variables, quasi-experiments, RCTs,R] # self-defined categories
# include-after-body: pub_ready_viz.html
citation: 
  url: https://bokola.github.io/posts/2026-01-15-survey-design/ 

draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
code-fold: true
# bibliography: reference.bib
---


# Multiple Regression

A number of assumptions underlying ordinary least square (OLS) regression come 
into close scrutiny under multiple linear regressions. Key assumptions include
**linearity, independence of residuals, homoscedasticity, normality of residuals**.

```{r libs}
pacman::p_load(
  AER, tidyverse, mvtnorm, flextable, modelsummary, sandwich, stargazer,
  MASS, rddtools, scales, broom, prettyunits
)
data("CASchools", package = "AER")

#' Render model coefficients as a flextable
#'
#' @param x model object
#'
#' @returns
#' @export
#'
#' @examples
render_summary <- function(x){
  x %>% broom::tidy() %>%
    mutate(p.value = prettyunits::pretty_p_value(p.value)) %>%
    mutate(across(where(is.numeric), ~ round(., 3))) %>%
    flextable::flextable()
}

#' Render a list of models as flextable
#'
#' @param models list object of models
#' @param robust_se list object of robust standard errors
#'
#' @returns
#' @export
#'
#' @examples
model_tables <- function(models, robust_se) {
  modelsummary(
    models,
    vcov = robust_se,
    # Applies your robust SEs
    output = "flextable",
    # Outputs as a flextable object
    fmt = 3,
    # 3 decimal places
    stars = TRUE,
    # Significance stars (optional)
    # gof_map = gm1 # Clean up GOF statistics
    coef_omit = "^\\(Intercept\\)$"
    ,
    gof_omit = NULL
  ) |>
    autofit() |>
    theme_vanilla()   %>%         # Clean academic style
    # This command removes any remaining horizontal lines inside the body
    border_inner_h(part = "body", border = officer::fp_border(width = 0))
}
```

```{r}
#| label: fig-ols
#| fig-cap: "OLS assumptions visual diagnostic"
#|fig-height: 18
m <- lm(math ~ expenditure, data = CASchools)

par(mfrow = c(2, 2))
plot(m, col = "steelblue", pch = 20)
par(mfrow = c(1, 1))
```
* A plot of residuals vs. fitted values assesses linearity and homoscedasticity assumption. 
A random (white noise) scatter around 0 suggests the assumptions are met.
* A normal quantile-quantile (Q-Q) plot assesses the normality assumption of the residuals
with departure from the straight line suggesting departure from normality.
* A plot of standardized residuals against fitted values assesses homoscedasticity assumption.
Look for white noise around the horizontal line in the scale location plot.
* The residual vs. leverage plot assesses for influential outliers. Points far away from the 
center are considered influential.

A number of factors may make multiple OLS regression biased due to a violation
of any of the assumptions underlying OLS. Sources of these biases include:

**omitted variables, misspecification of the functional form, measurement errors,
missing data and sample selection, simultaneous causality bias**.

## omitted variables

There is no easy fix for addressing omitted variables as a source of bias. Available guidelines stress

1. specifying coefficient(s) of interest
2. Identifying important sources of omitted variable bias using available 
domain knowledge prior to model fitting so you end up with a a baseline model 
specification and a questionable list of other regressors
3. Using different model specification to test whether questionable regressors
have effects different from zero
4. Providing full disclosure of your results such as results of different model 
specifications in support of your argument

The other sources of bias can be assessed via simulations.

## Misspecification of the functional form of a regression function

Assume the underlying function is such that $Y_i = X_i^2$, but the model is
specified in the form $Y_i = \beta_0 + \beta_1X_i + \epsilon_i$.

```{r}
#| label: fig-ms-lm
#| fig-cap: "Model misspecification"
set.seed(as.numeric(Sys.time()))

X <- runif(100, -5, 5)
Y <- X^2 + rnorm(100)

ml <- lm(Y ~ X)

plot(
  X,
  Y,
  main = "Model misspecification",
  pch = 20,
  col = "steelblue"
  
)

# add the regression line
abline(
  ml,
  col = "grey",
  lwd = 2, 
  lty = 2
)

legend(
  "bottomright",
  bg = "transparent",
  cex = 0.8,
  lwd = 2,
  lty = 2,
  col = "grey",
  legend = "OLS line"
  
)

```

## Measurement error and errors-in-variable bias

Set in due to imprecise measurement of the independent variable and do not disappear
not even with a large sample size. Due to imprecise measurements we observe 
$\overset{\sim}{X}_i$ instead of $X_i$ the 
model under consideration becomes 

\begin{align*}
  Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + \underbrace{\beta_1 (X_i - \overset{\sim}{X}_i) + u_i}_{=v_i}. \\
  Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + v_i.
\end{align*}

where $\overset{\sim}{X}_i$ and the error term $v_i$ are correlated. The OLS would
thus be biased and inconsistent for $B_1$ with the strength of bias dependent on the
correlation between the observed regressor $\overset{\sim}{X}_i$ and the measurement error
$w_i = X_i - \overset{\sim}{X}_i$.
The classical measurement error model assumes that the measurement error, $w_i$, has zero mean and that it is uncorrelated with the variable, $X_i$, and the error term of the population regression model, $u_i$:

\begin{equation}
  \overset{\sim}{X}_i = X_i + w_i, \ \ \rho_{w_i,u_i}=0, \ \ \rho_{w_i,X_i}=0. 
\end{equation}
Then it holds that 
$$\begin{equation}
  \widehat{\beta}_1 \xrightarrow{p}{\frac{\sigma_{X}^2}{\sigma_{X}^2 + \sigma_{w}^2}} \beta_1, 
\end{equation}
$${\#eq-cmembias}
which implies inconsistency as $\sigma_{X}^2, \sigma_{w}^2 > 0$ such that the fraction in (@eq-cmembias) is smaller than $1$. Note that there are two extreme cases: 

1. If there is no measurement error, $\sigma_{w}^2=0$ such that $\widehat{\beta}_1 \xrightarrow{p}{\beta_1}$. 

2. If $\sigma_{w}^2 \gg \sigma_{X}^2$ we have $\widehat{\beta}_1 \xrightarrow{p}{0}$. This is the case if the measurement error is so large that there essentially is no information on $X$ in the data that can be used to estimate $\beta_1$.

The most obvious way to deal with errors-in-variables bias is to use an accurately measured $X$. If this is not possible, instrumental variables regression is an option. One might also deal with the issue by using a mathematical model of the measurement error and adjust the estimates appropriately: if it is plausible that the classical measurement error model applies and if there is information that can be used to estimate the ratio in equation (@eq-cmembias), one could compute an estimate that corrects for the downward bias.


```{r}
#| label: fig-me
#| fig-cap: "Measurement error"
dat <- data.frame(
  rmvnorm(1000, c(50, 100), 
          sigma = cbind(c(10, 5), c(5, 10))))

# set columns names
colnames(dat) <- c("X", "Y")

# estimate the model (without measurement error)
noerror_mod <- lm(Y ~ X, data = dat)

# estimate the model (with measurement error in X)
dat$X <- dat$X + rnorm(n = 1000, sd = sqrt(10))
error_mod <- lm(Y ~ X, data = dat)

# print estimated coefficients to console
noerror_mod$coefficients
#> (Intercept)           X 
#>  76.3002047   0.4755264
error_mod$coefficients
#> (Intercept)           X 
#>   87.276004    0.255212
#>   
# plot sample data
plot(dat$X, dat$Y, 
     pch = 20, 
     col = "steelblue",
     xlab = "X",
     ylab = "Y")

# add population regression function
abline(coef = c(75, 0.5), 
       col = "darkgreen",
       lwd  = 1.5)

# add estimated regression functions
abline(noerror_mod, 
       col = "purple",
       lwd  = 1.5)

abline(error_mod, 
       col = "darkred",
       lwd  = 1.5)

# add legend
legend("topleft",
       bg = "transparent",
       cex = 0.8,
       lty = 1,
       col = c("darkgreen", "purple", "darkred"), 
       legend = c("Population", "No Errors", "Errors"))
```

## Missing data and sample selection

If data is not missing at random, we are faced with biased estimates.

```{r}
#| label: fig-mis
#| fig-cap: "Missing data bias"
# randmly select 500 obs
id <- sample(1:1000, 500)
dat1 <- dat[-id,]
dat2 <- dat[id,]

ci_mod <- lm(Y ~ X, data = dat1)

plot(
  dat1$X,
  dat1$Y,
  col = "steelblue",
  pch = 20,
  cex = 0.8,
  xlab = "X",
  ylab = "Y"
)

points(
  dat2$X,
  dat2$Y,
  cex = 0.8,
  col = "grey",
  pch = 20
)

# add pop reg function
abline(
  coef = c(75, 100),
  col = "darkgreen",
  lwd = 1.5
)

# add full sample model
abline(noerror_mod, col = "black")

# add model without a sample of 500
abline(ci_mod, col = "purple")

legend(
  "bottomright",
  lty = 2,
  bg = "transparent",
  cex = 0.8,
  col = c("darkgreen", "black", "purple"),
  legend = c("population", "full sample", "random sample of 500")
)
```

## Simultaneous causality bias

Arises when `Y` influences `X`. Let's assume that we are interested in estimating the
effect of 20% increase in price on cigarette. We get the log of price and use it to
model consumption, giving a price elasticity of consumption interpretation. Since
it is not demand on the Y-axis, this not an estimate of the demand curve, and is 
therefore a classic example of reverse (simultaneous) causality.

```{r}
#| label: fig-revc
#| fig-cap: "Reverse causlaity bias"
data("CigarettesSW")
c1995 <- subset(CigarettesSW, year == "1995")

# estimate the model
cigcon_mod <- lm(log(packs) ~ log(price), data = c1995)
cigcon_mod

# plot the estimated regression line and the data
plot(log(c1995$price), log(c1995$packs),
     xlab = "ln(Price)",
     ylab = "ln(Consumption)",
     main = "Demand for Cigarettes",
     pch = 20,
     col = "steelblue")

abline(cigcon_mod, 
       col = "darkred", 
       lwd = 1.5)
# add legend
legend("topright",lty=1,col= "darkred", "Estimated Regression Line")
```

# Instrumental variable regression

## Simple instrumental variable regression

Instrumental variables (IV) can be used to address possible omitted variable bias.
Considering a simple regression model 

$$
\begin{equation}
Y_i = \beta_0 + \beta_1X_i + \epsilon_i, i = 1, \ldots, n
\end{equation}
$${\#eq-lm}

in which the error term is correlated with the endogenous regressor $X$ making the OLS estimate
inconsistent for $\beta_1$, we can use a single IV variable $Z$ to obtain a 
consistent estimator for $\beta_1$. The IV variable has to meet the 


1. **relevance condition**: $X$ and its instrument $Z$ must be correlated
$\rho_{Z_i, X_i} \ne 0$ or multicollinear for multiple IV regression.
2. **instrument exogeneity condition**: The instrument $Z$ must not be 
correlated with the error term. 
$\rho_{Z_i, \epsilon_i} = 0$

Estimation is through a two-stage least square (TSLS) estimation. The first stage
condenses the variation in the endogenous variable $X$ into a problem-free part 
that is explained by the IV variable $Z$, and a problematic component that is 
correlated to the error term $\epsilon_i$. In the second stage the problem-free
component of the variation in $X$ is used to estimate $\beta_1$.

The first stage uses the regression model $X_i = \pi_0 + \pi_1Z_i + v_i$, where 
$\pi_0 + \pi_1Z_i$ is the component of $X_i$ explained by $Z_i$, and $v_i$ is the 
component that can not be explained by $Z_i$ and is correlated with $\epsilon_i$.
We can use OLS estimates $\hat{\pi_0}$ and $\hat{\pi_1}$ to predict $X_i$ giving 
problem-free $\hat{X_i}$ ($\hat{X_i}$ exogenous in the regression of Y on $\hat{X}$
done on second stage ) if $Z$ is indeed a valid instrument. The second stage 
produces $\hat{\beta}_0^{TSLS}$, and $\hat{\beta}_1^{TSLS}$, which are the TSLS 
estimates of $\beta_0$ and $\beta_1$ respectively. For a single IV it can be shown 
that the TSLS estimator is but the ratio of sample covariance between $Z$ and $Y$

$$
\begin{equation}
\hat{\beta}^{TSLS} = \frac{s_{Z, Y}}{s_{Z, X}} = \frac{\frac{1}{n-1} \sum_{i=1}^n (Y_i - \bar{Y})(Z_i - \bar{Z})}{\frac{1}{n-1}\sum_{i=1}^n (X_i - \bar{X}) (Z_i - \bar{Z})}
\end{equation}
$${\#eq-tsls}

### Example: Demand for cigarettes

```{r}
skimr::skim(CigarettesSW)
```

We wish to estimate $\beta_1$ in 

$$
\begin{equation}
log(Q_i) = \beta_0 + \beta_1 log(P_i) + \epsilon_i
\end{equation}
$${\#eq-tcigar}


where $Q_i$ is the number of cigarette packs per capita sold and $P_i$ is the 
after tax average real price per pack of cigarette in state $i$. We use sales tax
(dollars per pack) to instrument the endogenous regressor $log(P_i)$. Salestax is
considered a relevant instrument as it is included in the after-tax average 
price per pack.

```{r}
# compute real per capita prices
CigarettesSW$rprice <- with(CigarettesSW, price / cpi)

#  compute the sales tax
CigarettesSW$salestax <- with(CigarettesSW, (taxs - tax) / cpi)

# check the correlation between sales tax and price
cor(CigarettesSW$salestax, CigarettesSW$price)
#> [1] 0.6141228

# generate a subset for the year 1995
c1995 <- subset(CigarettesSW, year == "1995")
```

The first stage regression is $log(P_i) = \pi_0 + \pi_1SalesTax_i + v_i$. In the
second stage we regress $log(Q_i)$ on $log(P_i)$ to obtain $\hat{\beta_0}^{TSLS}$
$\hat{\beta_1}^{TSLS}$.

```{r}
# perform the first stage regression
cig_s1 <- lm(log(rprice) ~ salestax, data = c1995)

coeftest(cig_s1, vcov = vcovHC, type = "HC1") %>% broom::tidy() %>%
  mutate(p.value = prettyunits::pretty_p_value(p.value)) %>% 
  flextable::flextable()

# store the fitted values from first stage
lcigp_pred <- cig_s1$fitted.values

```

Run the second stage:

```{r}
cig_s2 <- lm(log(c1995$packs) ~ lcigp_pred)
coeftest(cig_s2, vcov. = vcovHC) %>% broom::tidy() %>%
  mutate(p.value = prettyunits::pretty_p_value(p.value)) %>%
  flextable::flextable()

```

Thus estimating the model (@eq-tcigar) using TSLS yields 

$$
\begin{equation}
  \widehat{\log(Q_i)} = \underset{(1.70)}{9.72} - \underset{(0.36)}{1.08} \log(P_i)
\end{equation}
$${\#eq-ecigstsls}

where we write $\log(P_i)$ instead of $\widehat{\log(P_i)}$ for consistency. See
`AER::ivreg()` for automatic IV regression.

## The general IV regression model

An extension of simple IV regression to multiple predictors: 

$$
\begin{equation}
  Y_i = \beta_0 + \beta_1 X_{1i} + \dots + \beta_k X_{ki} + \beta_{k+1} W_{1i} + \dots + \beta_{k+r} W_{ri} + \epsilon_i, 
\end{equation}
$${\#eq-givmodel}

with $i=1,\dots,n$ is the general instrumental variables regression model where

- $Y_i$ is the dependent variable,

- $\\beta_0,\dots,\beta_{k+1}$ are $1+k+r$ unknown regression coefficients,

- $X_{1i},\dots,X_{ki}$ are $k$ endogenous regressors ,

- $W_{1i},\dots,W_{ri}$ are $r$ exogenous regressors which are uncorrelated with $u_i$,

- $\epsilon_i$ is the error term,

- $Z_{1i},\dots,Z_{mi}$ are $m$ instrumental variables.

The coefficients are overidentified if $m>k$. If $m<k$, the coefficients are underidentified and when $m=k$ they are exactly identified. For estimation of the IV regression model we require exact identification or overidentification. The computations can be autometed with `AER::ivreg()` function.

Suppose we wish to estimate the model $Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + W_{1i} + \epsilon_i$,
where $X_{1i}$ and $X_{2i}$ are endogenous variables instrumented by $Z_{1i}$ and $Z_{2i}$ respectively, and $W_{1i}$ is an exogenous variable, a correct model specification involves listing all exogenous variables as
instruments too by joining them with `+`s on the right of the vertical bar `|`:

`y ~ X1 + X2 + W1 | W1 + Z1 + Z2`. A convenient way of specifying a model is
updating a formula using `.` (which includes all variables except for the outcome).
For instance, if we have one exogenous regressor $w1$
and one endogenous regrossor $x1$ with instrument $z1$, the appropriate formula
would be `y ~ w1 + x1 | w1 + z1` $=$ `y ~ w1 + x1 | . ~ x1 + z1`.

Some assumptions underlying the IV regression are:

1. $E(\epsilon_i|W_{1i}, \dots, W_{ri}) = 0$
2. $(X_{1i}, \dots, X_{ki}, W_{1i}, \dots, W_{ri}, Z_{1i}, \dots, Z_{mi})$ are i.i.d draws from their joint distribution
3. All variables have nonzero finite fourth moment, i.e., outliers are unlikely
4. The $Z_s$ are valid instruments.

### Example: Demands for cigarettes

We update (@eq-tcigar) to include income:


$$
\begin{equation}
log(Q_i) = \beta_0 + \beta_1 log(P_i) + \beta_2 log(income_i) +  \epsilon_i
\end{equation}
$${\#eq-ticigar}

where income is defined as real per capita income.

```{r}
# add rincome to the dataset
CigarettesSW$rincome <- with(CigarettesSW, income / population / cpi)

c1995 <- subset(CigarettesSW, year == "1995")

# estimate the model
cig_ivreg2 <- ivreg(log(packs) ~ log(rprice) + log(rincome) | log(rincome) + 
                    salestax, data = c1995)

coeftest(cig_ivreg2, vcov = vcovHC, type = "HC1") %>%
  broom::tidy() %>%
  mutate(p.value = prettyunits::pretty_p_value(p.value)) %>%
  flextable::flextable()


```

We obtain 

$$
\begin{equation}
  \widehat{\log(Q_i)} = \underset{(1.26)}{9.43} - \underset{(0.37)}{1.14} \log(P_i) + \underset{(0.31)}{0.21} \log(income_i)
\end{equation}
$${\#eq-eicigstsls}


Adding salestax as an IV:

```{r}
# add cigtax to the data set
CigarettesSW$cigtax <- with(CigarettesSW, tax/cpi)

c1995 <- subset(CigarettesSW, year == "1995")

# estimate the model
cig_ivreg3 <- ivreg(log(packs) ~ log(rprice) + log(rincome) | 
                    log(rincome) + salestax + cigtax, data = c1995)

coeftest(cig_ivreg3, vcov = vcovHC, type = "HC1") %>% 
  broom::tidy() %>%
  mutate(p.value = prettyunits::pretty_p_value(p.value)) %>%
  flextable::flextable()

```

$$
\begin{equation}
  \widehat{\log(Q_i)} = \underset{(0.96)}{9.89} - \underset{(0.25)}{1.28} \log(P_i) + \underset{(0.25)}{0.28} \log(income_i)
\end{equation}
$${\#eq-eicigstsls-2}

A question then arises on which estimates to trust. One could argue that the 
estimates from (@eq-eicigstsls-2) are more trustworthy due to the lower standard
errors compared to (@eq-eicigstsls). This is the subject considered under instrument 
validity.

## Checking instrument validity

1. **Weak instrument**: To check for a weak instrument, compute the $F$-statistic
corresponding to the hypothesis $H_0: Z_1 = \dots Z_m = 0$. A rule of thumb is
that an $F$-statistic less than 10 imply weak instruments. If your instruments
are weak, you can discard them and find better instruments or stick with them
but find methods that improve TSLS.
2. **Instrument exogeneity assumption is not met**: When we observe a correlation
between an instrument and the error term, the IV regression is not consistent. 
Overidentifying restriction test (J-test) is used to test that additional instruments
are exogenous. It follows $\chi^2_{m -k}$ (under homoscedasticity), where $m$ is the number of instruments
and $k$ the number of endogenous regressors. It involves regressing the error term
on all instruments and endogenous variables:

$$
\begin{equation}
\hat{\epsilon}^{TSLS} = \delta_0 + \delta_1Z_{1i} + \dots + \delta_mZ_{mi} +
\delta_{m+1}W_{1i} + \dots + \delta_{m+r}W_{ri} + e_i
\end{equation}
$${\#eq-jstat}

and testing whether all instruments are exogenous through the joint hypothesis
$H_0: \delta_1 = 0, \dots, \delta_m = 0$, and computing the statistic $J = mF$.
(see <https://www.econometrics-with-r.org/12.4-attdfc.html>)


### Example: Demand for cigarettes

We assess whether general sales tax and the cigarette-specific tax are valid
instruments. It could be argued that tobacco growing states find economic importance
in lower tobacco taxes and may lobby for them staying low. It may also be possible
that tobacco growing states have higher rates of smoking than others, which would 
lead to endogeneity of cigarette-specific taxes. This can only be solved by including
data on the size of the tobacco and cigarette industry into the regression, which is
not available in this case.

It can be assumed that the role of tobacco and cigarette industry differs across
states, but not across time. We can then exploit the panel structure of the data
at two different time periods (1985, 1995) to estimate the *long-run elasticity*
of the demand for cigarettes:

$$
\begin{equation}
log(Q_{i, 1995}) - log(Q_{i, 1985}) = \beta_0 + \beta_1[log(P_{i, 1995}) -
log(P_{i, 1985})] + \beta_2[log(income_{i, 1995}) - log(income_{i, 1985})] + \epsilon_i
\end{equation}
$${\#eq-elas}


```{r}
# subset data for year 1985
c1985 <- subset(CigarettesSW, year == "1985")

# define differences in variables
packsdiff <- log(c1995$packs) - log(c1985$packs)

pricediff <- log(c1995$price/c1995$cpi) - log(c1985$price/c1985$cpi)

incomediff <- log(c1995$income/c1995$population/c1995$cpi) -
log(c1985$income/c1985$population/c1985$cpi)

salestaxdiff <- (c1995$taxs - c1995$tax)/c1995$cpi - 
                                  (c1985$taxs - c1985$tax)/c1985$cpi

cigtaxdiff <- c1995$tax/c1995$cpi - c1985$tax/c1985$cpi
```

We then perform 3 different IV estimations using `ivreg()`for differences between
1985, 1995:

1. TSLS using only the difference in sales taxes as the IV 
2. TSLS using only difference in cigarette-specific taxes
3. TSLS using both differences in sales taxes and cigarette-specific taxes

```{r}
# estimating the 3 models
cig_ivreg_diff1 <- ivreg(
  packsdiff ~ pricediff + incomediff | incomediff + salestaxdiff
)

cig_ivreg_diff2 <- cig_ivreg_diff1 <- ivreg(
  packsdiff ~ pricediff + incomediff | incomediff + cigtaxdiff
)

cig_ivreg_diff3 <- cig_ivreg_diff1 <- ivreg(
  packsdiff ~ pricediff + incomediff | incomediff + salestaxdiff + cigtaxdiff
)
```

Sales tax as IV

```{r}
#| label: tbl-sltx
#| tbl-cap: "Difference in sales tax as IV"

coeftest(cig_ivreg_diff1, vcov. = vcovHC, type = "HC1") %>%
  render_summary()

```

Cigarette-specific tax as IV

```{r}
#| label: tbl-cigtx
#| tbl-cap: "Cigarette-specific tax as IV"

coeftest(cig_ivreg_diff2, vcov. = vcovHC, type = "HC1") %>% 
  render_summary()
```

Both sales tax and cigarette-specific tax as IV

```{r}
#| label: tbl-both
#| tbl-cap: "Both sales tax and cigarette-specific tax as IV"

coeftest(cig_ivreg_diff3, vcov. = vcovHC, type = "HC1") %>%
  render_summary()
```

A tabulated summary

```{r}
#| label: tbl-iv-comparison
#| tbl-cap: "Dependent Variable: 1985-1995 Difference in Log per Pack Price"
#| message: false
#| warning: false

# 1. Define the models in a named list (this replaces column.labels)
models <- list(
  "IV: salestax" = cig_ivreg_diff1,
  "IV: cigtax"   = cig_ivreg_diff2,
  "IVs: salestax, cigtax" = cig_ivreg_diff3
)

# Define which Goodness-of-Fit stats to show and what to call them
gm <- list(
  list("raw" = "nobs", "clean" = "Observations", "fmt" = 0),
  list("raw" = "r.squared", "clean" = "R2", "fmt" = 3),
  list("raw" = "adj.r.squared", "clean" = "Adj. R2", "fmt" = 3)
)

# 2. Define standard errors
# Note: modelsummary can calculate HC1 automatically, 
# but since you already have 'rob_se', we can pass it directly.
rob_se_list <- list(
  sqrt(diag(vcovHC(cig_ivreg_diff1, type = "HC1"))),
  sqrt(diag(vcovHC(cig_ivreg_diff2, type = "HC1"))),
  sqrt(diag(vcovHC(cig_ivreg_diff3, type = "HC1")))
)

# 3. Generate the table
modelsummary(
  models,
  vcov = rob_se_list,        # Applies your robust SEs
  output = "flextable",      # Outputs as a flextable object
  fmt = 3,                   # 3 decimal places
  stars = TRUE,              # Significance stars (optional)
  gof_map = gm # Clean up GOF statistics
) |> 
  autofit() |> 
  theme_vanilla()   %>%         # Clean academic style
# This command removes any remaining horizontal lines inside the body
  border_inner_h(part = "body", border = officer::fp_border(width = 0))


```

We observe negative estimates for `pricediff` that differ when `cigtax` is used 
as IV. We proceed to test the validity of the instruments by computing F-statistics
for the first-stage regressions.

```{r}
#| label: tbl-ivrel
#| tbl-cap: "Test IV relevance by an F-test. All the F-statistics are significantly different from zero"
# first-stage regressions
mod_relevance1 <- lm(pricediff ~ salestaxdiff + incomediff)
mod_relevance2 <- lm(pricediff ~ cigtaxdiff + incomediff)
mod_relevance3 <- lm(pricediff ~ incomediff + salestaxdiff + cigtaxdiff)

# check instrument relevance

mods <- list(mod_relevance1, mod_relevance2, mod_relevance3)
names(mods) <- c("salestax", "cigartax", "both")
hyp <- list("salestaxdiff = 0", "cigtaxdiff = 0", c("salestaxdiff = 0", "cigtaxdiff = 0"))

iv_tests <- map2(
  mods, hyp, \(m, h){
    linearHypothesis(m, h, vcov. = vcovHC(m, type = "HC1"))
  }
) 

 iv_tests |>
  map_dfr(broom::tidy, .id = "model") %>% 
   mutate(p.value = prettyunits::pretty_p_value(p.value)) %>%
   mutate(
     across(where(is.numeric), ~ round(., 3))
   ) %>% 
   flextable()

```

Next we conduct overidentifying restriction test for model with both sales tax and 
cigatax as IV which is the only model where the coefficient on the difference in
log prices is overidentified ($m = 2, k=1$) making it possible to compute the $J$
statistic. To achive this, we regress the residuals from `cig_ivreg_diff3` on the
IVs and the presumed exogenous regressor `incomediff`. We again test if the coefficients
of the IV variables is zero, a condition necessary for exogenous variables.

```{r}
#| label: tbl-jtest
#| tbl-cap: "J-test statitic testing for exogeneity condition of IVs"
# compute the J-statistic
cig_iv_OR <- lm(residuals(cig_ivreg_diff3) ~ incomediff + salestaxdiff + cigtaxdiff)

cig_OR_test <- linearHypothesis(cig_iv_OR, 
                               c("salestaxdiff = 0", "cigtaxdiff = 0"), 
                               test = "Chisq")
cig_OR_test %>% render_summary()

```

**Note**: The reported degrees of freedom is wrong, (2 instead of $m-k = 2-1=1$).
The J-statistic is $\chi^2_1$ instead of $\chi^2_2$ as reported by the model.
We calculate the correct p-value:

```{r}
pchisq(cig_OR_test[2, 5], df = 1, lower.tail = FALSE)
```

We reject the hypothesis that both instruments are exogenous at $\alpha = 0.05$
and conclude that

1. sales tax is an invalid instrument for the per-pack price,
2. and so is the cigarette-specific tax,
3. and that both instruments are invalid.

# Experiments and Quasi-experiments

Experiments give us the tools to conduct program evaluation interest being to 
measure causal effects of programs. The gold-standard for this kind of work is 
a randomized controlled experiment. In a quasi-experiment there are no randomized
assignments to interventions, we study the processes in their natural setting.

## Potential outcomes, causal effects and idealized experiments

Interest is in the **average causal effect** estimated by the **difference estimator**.
A potential outcome is the outcome for an individual under a potential treatment.
For this individual the causal effect of the treatment is the difference between 
the potential outcome if the individual receives the treatment and the potential
outcome if they do not. We are interested in studying the **average causal effect**
since this difference may vary from individual to individual. In a randomized 
controlled experiment, it is expected that
1. Subjects are randomly selected from the population(ensures average causal effect
in sample equals average causal effect in population),
2. Subjects are randomly assigned to treatment and control group (ensures receiving
a treatment is independent of a subject's potential outcome).

If above conditions are met, then the average causal effect is the expected outcome
in treatment group less expected outcome in the control:

$$
\text{Average causal effect}  = E(Y_i|X_i = 1) - E(Y_i|X_i = 0)
$$

where $X_i$ is a binary treatment indicator. Estimation follows **differences estimator** 
which is an OLS regression model

$$
\begin{equation}
Y_i = \beta_0 + \beta_1X_i + \epsilon_i, i = 1, \dots, n
\end{equation}
$${\#eq-ae}

where random assignment ensures $E(\epsilon_i|X_i) = 0$. The difference estimator
in (@eq-ae) can take additional regressors $W_1, \dots, W_r$ and is called a
**differences estimator with additional regressors**:

$$
\begin{equation}
Y_i = \beta_0 + \beta_1X_i + \beta_2W_{1i} + \dots + \beta_{1+r}W_{ri} +
\epsilon_i, i = 1, \dots, n
\end{equation}
$${\#eq-ae-dear}

It utilizes a **conditional mean independence** assumption that states that
treatment assignment $X_i$ is random and independent of any pretreatment variable
$W_i$. This ensures that

$$
\begin{equation}
E(\epsilon_i|X_i, W_i) = E(\epsilon_i|W_i) = 0
\end{equation}
$$
 
 implying that the conditional expectation of the error term $\epsilon_i$, given 
 the treatment indicator $X_i$ and the the pretreatment characteristic $W_i$ does
 not depend on $X_i$.
 
 ### Analysis of experimental data
 
 `STAR` data set from `{AER}` is a large student-teacher achievement ratio randomized 
 controlled experiment that was assessing whether class size reduction is effective
 in improving education outcomes. 6400 students were randomly assigned into one
 of the three interventions: small class (13 to 17 students per teacher),
 regular class (22 to 25 students per teacher), and regular-with-aide class
 (22 to 25 students with a full-time teacher’s aide). Teachers were also randomly
 assigned to classes. The subjects were followed from kindergarten to grade 3 
 (K,1, 2, 3 suffixes in variable names).
 
```{r}
data("STAR", package = "AER")
skimr::skim(STAR)
```
 
 In essence, there were 2 treatments in each grade (small class with 13 - 17
 students, and a regular class with 22 - 25 students and a teaching aide). We
 introduce two binary variables, each being an indicator for the respective treatment
 group for the difference estimator. This is to capture the treatment effect for each treatment 
 group separately yielding the population regression model for the test score as
 
 $$
 \begin{equation}
 Y_i = \beta_0 + \beta_1 smallClass_i + \beta2RegAide_i + \epsilon_i
 \end{equation}
 $${\#eq-pop-model}
 
```{r}
#| label: tbl-popmod1
#| tbl-cap: "Project STAR: Differences estimate. Class size reduction improves study performance excpt for grade 1"
#| tbl-cap-location: bottom
grades <- c("k",1:3)

# map over grades to create a list of models

star_models <- grades |>
  set_names(paste0("grade_", grades)) |>
  map(
    \(y){
#       construct the formula dynamically: I(ready + mathy) ~ starY
      form <- as.formula(paste0("I(read", y, "+ math", y, ") ~ star", y))
      lm(form, data = STAR)
    }
  )

# compute robust standard error
rob_se <- star_models |>
 map(\(m) sqrt(diag(vcovHC(m, type = "HC1"))))

# Define which Goodness-of-Fit stats to show and what to call them
gm1 <- list(
  list("raw" = "nobs", "clean" = "Observations", "fmt" = 0),
  list("raw" = "r.squared", "clean" = "R2", "fmt" = 3),
  list("raw" = "adj.r.squared", "clean" = "Adj. R2", "fmt" = 3),
  # Added Residual Standard Error
  list("raw" = "sigma", "clean" = "Residual Std. Error", "fmt" = 3),
  # Added F-statistic
  list("raw" = "statistic", "clean" = "F-statistic", "fmt" = 3)
)

# render table

model_tables(models = star_models, robust_se = rob_se)
  
```
 
We can augment the model in (@eq-pop-model) with extra variables under 2 
circumstances: 

1. Adding the regressors explain some of the observed variations in the outcome
making the estimates more efficient
2. If treatment assignment is not random, the estimates obtained in (@eq-pop-model)
are possibly biased. The additional regressors could mitigate this challenge.

For our setting we could consider 6 additional variables (in a staggered manner):

* *experience* - Teacher's years of experience
* *boy* - A dummy (student is a boy)
* *lunch* - Free lunch eligibility (dummy)
* *black* - Student is black (dummy)
* *race* - Student's race is other than black or white (dummy)
* *schoolid* - School indicator variables

**Note**: For randomized controlled experiments, we can make causal claims only 
for variables that subjects were randomly assigned into.

## Quasi experiments

Quasi experiments exploit "as if" randomness to use methods considered under 
randomized controlled experiments. They come in two forms:

1. Random variations in individual circumstances allow to view the treatment 
"as if" it was randomly determined.
2. The treatment is only partly determined by "as if" random variation.

The former allows the use of either model (@eq-ae-dear), i.e.,
**the difference estimator with additional regressors**, or, using 
**difference-in-difference (DID)** if the "as if" randomness
does not guarantee the non-existence of systematic differences in the control
and treatment groups. The latter case invites the use of an IV approach for a
case like model (@eq-ae-dear), utilizing the source of "as if" randomness in 
treatment assignment as the instrument. Other methods include 
**sharp regression discontinuity design (RDD)** and 
**fuzzy regression discontinuity design (FRDD)**. We use simulated data
to explore these methods

### The difference-in-difference (DID) estimator

The source of "as if" randomness in treatment assignment in quasi-experiments does
not always rule out systematic differences between control and treatment groups.
DID involves data in organised in the before and after treatment. 

$$
\begin{align}
  \widehat{\beta}_1^{\text{diffs-in-diffs}} =& \, (\overline{Y}^{\text{treatment,after}} - \overline{Y}^{\text{treatment,before}}) - (\overline{Y}^{\text{control,after}} - \overline{Y}^{\text{control,before}}) \\
  =& \Delta \overline{Y}^{\text{treatment}} - \Delta \overline{Y}^{\text{control}}, 
\end{align}
$${\#eq-DID}

with

+ $\overline{Y}^{\text{treatment,before}}$ - the sample average in the treatment group before the treatment

+ $\overline{Y}^{\text{treatment,after}}$ - the sample average in the treatment group after the treatment

+ $\overline{Y}^{\text{treatment,before}}$ - the sample average in the control group before the treatment

+ $\overline{Y}^{\text{treatment,after}}$ - the sample average in the control group after the treatment.

We now use R to plot a DID setting.

```{r}
#| label: fig-did
#| fig-cap: "Difference-in-differnce plot"
# initialize plot and add control group
plot(c(0, 1), c(6, 8), 
     type = "p",
     ylim = c(5, 12),
     xlim = c(-0.3, 1.3),
     # main = "The Differences-in-Differences Estimator",
     xlab = "Period",
     ylab = "Y",
     col = "steelblue",
     pch = 20,
     xaxt = "n",
     yaxt = "n")

axis(1, at = c(0, 1), labels = c("before", "after"))
axis(2, at = c(0, 13))

# add treatment group
points(c(0, 1, 1), c(7, 9, 11), 
       col = "darkred",
       pch = 20)

# add line segments
lines(c(0, 1), c(7, 11), col = "darkred")
lines(c(0, 1), c(6, 8), col = "steelblue")
lines(c(0, 1), c(7, 9), col = "darkred", lty = 2)
lines(c(1, 1), c(9, 11), col = "black", lty = 2, lwd = 2)

# add annotations
text(1, 10, expression(hat(beta)[1]^{DID}), cex = 0.8, pos = 4)
text(0, 5.5, "s. mean control", cex = 0.8 , pos = 4)
text(0, 6.8, "s. mean treatment", cex = 0.8 , pos = 4)
text(1, 7.9, "s. mean control", cex = 0.8 , pos = 4)
text(1, 11.1, "s. mean treatment", cex = 0.8 , pos = 4)
```

The DID estimator (@eq-DID) can take a regression notation: $\hat{\beta_1}^{DID}$
is the OLS estimator of $\beta_1$ in 

$$
\begin{equation}
\Delta Y_i = \beta_0 + \beta_1X_i + \epsilon_i
\end{equation}
$${\#eq-DID-1}

with $\Delta Y_i$ denoting the difference in pre and post-treatment outcomes of 
subject $i$, $X_i$ is the treatment indicator. We can add additional regressors measuring
pretreatment characteristics to (@eq-DID-1) to get the 
**difference-in-difference estimator with additional regressors**.


$$
\begin{equation}
\Delta Y_i = \beta_0 + \beta_1X_i + \beta_2W_{1i} + \dots + \beta_{1+r}W_{ri} +  \epsilon_i
\end{equation}
$${\#eq-DID-2}

Simulate data:

```{r}
set.seed(as.numeric(Sys.Date()))
# set sample size
n <- 200

# define treatment effect
TEffect <- 4

# generate treatment dummy
TDummy <- c(rep(0, n/2), rep(1, n/2))

# simulate pre- and post-treatment values of the dependent variable
y_pre <- 7 + rnorm(n)
y_pre[1:n/2] <- y_pre[1:n/2] - 1
y_post <- 7 + 2 + TEffect * TDummy + rnorm(n)
y_post[1:n/2] <- y_post[1:n/2] - 1 
```

Plot data:

```{r}
#| label: fig-did-1
#| fig-cap: "Artificial data for DID estimation"
pre <- rep(0, length(y_pre[TDummy==0]))
post <- rep(1, length(y_pre[TDummy==0]))

# plot control group in t=1
plot(jitter(pre, 0.6), 
     y_pre[TDummy == 0], 
     ylim = c(0, 16), 
     col = alpha("steelblue", 0.3),
     pch = 20, 
     xlim = c(-0.5, 1.5),
     ylab = "Y",
     xlab = "Period",
     xaxt = "n"
     # ,main = "Artificial Data for DID Estimation"
     )

axis(1, at = c(0, 1), labels = c("before", "after"))

# add treatment group in t=1
points(jitter(pre, 0.6), 
       y_pre[TDummy == 1], 
       col = alpha("darkred", 0.3), 
       pch = 20)

# add control group in t=2
points(jitter(post, 0.6),
       y_post[TDummy == 0], 
       col = alpha("steelblue", 0.5),
       pch = 20)

# add treatment group in t=2
points(jitter(post, 0.6), 
       y_post[TDummy == 1], 
       col = alpha("darkred", 0.5),
       pch = 20)
```

Compute DID estimate by hand:

```{r}
# compute the DID estimator for the treatment effect 'by hand'
mean(y_post[TDummy == 1]) - mean(y_pre[TDummy == 1]) - 
(mean(y_post[TDummy == 0]) - mean(y_pre[TDummy == 0]))

```

Compute DID using a linear model:

```{r}
# compute the DID estimator using a linear model
lm(I(y_post - y_pre) ~ TDummy) %>% render_summary()

```

### Regression discontinuity estimators

Consider the model

$$
\begin{align}
  Y_i =& \beta_0 + \beta_1 X_i + \beta_2 W_i + u_i
\end{align}
$${\#eq-SRDDsetting}

and let

$$
\begin{align*}
X_i =& 
  \begin{cases}
    1, & W_i \geq c \\
    0, & W_i < c,
  \end{cases}
\end{align*}
$$
so that the receipt of treatment, $X_i$, is determined by some threshold $c$ of a continuous variable $W_i$, the so called running variable. The idea of *regression discontinuity design* is to use observations with a $W_i$ close to $c$ for estimation of $\beta_1$. $\beta_1$ is the average treatment effect for individuals with $W_i = c$ which is assumed to be a good approximation to the treatment effect in the population. (@eq-SRDDsetting) is called a *sharp regression discontinuity design* because treatment assignment is deterministic and discontinuous at the cutoff: all observations with $W_i < c$ do not receive treatment and all observations where $W_i \geq c$ are treated.  

```{r}
#| label: fig-rdd
#| fig-cap: "Regression discontinuity plot"
set.seed(as.numeric(Sys.Date()))

# generate some sample data
W <- runif(1000, -1, 1)
y <- 3 + 2 * W + 10 * (W>=0) + rnorm(1000)

# construct rdd_data 
data <- rdd_data(y, W, cutpoint = 0)

# plot the sample data
plot(data,
     col = "steelblue",
     cex = 0.35, 
     xlab = "W", 
     ylab = "Y")
```

The argument `nbins` sets the number of bins the running variable is divided 
into for aggregation. The dots represent bin averages of the outcome variable.

We use the function `rdd_reg_lm()` to estimate the treatment effect using model
(@eq-SRDDsetting). `slope = "same"` restricts the slopes of the estimated regression
function to be the same on both sides of the jump at cutpoint $W = 0$.

```{r}
rdd_mod <- rdd_reg_lm(rdd_object = data, 
                      slope = "same")
render_summary(rdd_mod)
```

You can visualize by calling plot

```{r}
# plot the RDD model along with binned observations
plot(rdd_mod,
     cex = 0.35, 
     col = "steelblue", 
     xlab = "W", 
     ylab = "Y")
```

So far we assumed that crossing of the threshold determines receipt of treatment so that the jump of the population regression functions at the threshold can be regarded as the causal effect of the treatment.

When crossing the threshold $c$ is not the only cause for receipt of the treatment, treatment is no a deterministic function of $W_i$. Instead, it is useful to think of $c$ as a threshold where the *probability* of receiving the treatment jumps.

This jump may be due to unobservable variables that have impact on the probability of being treated. Thus, $X_i$ in (@eq-SRDDsetting) will be correlated with the error $u_i$ and it becomes more difficult to consistently estimate the treatment effect. In this setting, using a *fuzzy regression discontinuity design* which is based an IV approach may be a remedy: take the binary variable $Z_i$ as an indicator for crossing of the threshold,

$$
\begin{align*}
  Z_i = \begin{cases}
    1, & W_i \geq c \\
    0, & W_i < c,
  \end{cases}
\end{align*}
$$
and assume that $Z_i$ relates to $Y_i$ only through the treatment indicator $X_i$. Then $Z_i$ and $u_i$ are uncorrelated but $Z_i$ influences receipt of treatment so it is correlated with $X_i$. Thus, $Z_i$ is a valid instrument for $X_i$ and (@eq-SRDDsetting) can be estimated using TSLS.  

The following code chunk generates sample data where observations with a value of the running variable $W_i$ below the cutoff $c=0$ do not receive treatment and observations with $W_i \geq 0$ do receive treatment with a probability of $80\%$ so that treatment status is only partially determined by the running variable and the cutoff. Treatment leads to an increase in $Y$ by $2$ units. Observations with $W_i \geq 0$ that do not receive treatment are called *no-shows*: think of an individual that was assigned to receive the treatment but somehow manages to avoid it.


```{r}
# generate sample data
set.seed(as.numeric(Sys.Date()))
mu <- c(0, 0)
sigma <- matrix(c(1, 0.7, 0.7, 1), ncol = 2)

set.seed(1234)
d <- as.data.frame(mvrnorm(2000, mu, sigma))
colnames(d) <- c("W", "Y")

# introduce fuzziness
d$treatProb <- ifelse(d$W < 0, 0, 0.8)

fuzz <- sapply(X = d$treatProb, FUN = function(x) rbinom(1, 1, prob = x))

# treatment effect
d$Y <- d$Y + fuzz * 2
```

Plot:

```{r}
# generate a colored plot of treatment and control group
plot(d$W, d$Y,
     col = c("steelblue", "darkred")[factor(fuzz)], 
     pch= 20, 
     cex = 0.5,
     xlim = c(-3, 3),
     ylim = c(-3.5, 5),
     xlab = "W",
     ylab = "Y")

# add a dashed vertical line at cutoff
abline(v = 0, lty = 2)
#add legend
legend("topleft",pch=20,col=c("steelblue","darkred"),
       legend=c("Do not receive treatment","Receive treatment"))
```

Obviously, receipt of treatment is no longer a deterministic function of the running variable $W$. Some observations with $W\geq0$ *did not* receive the treatment. We may estimate a FRDD by additionally setting `treatProb` as the assignment variable `z` in `rdd_data`. Then `rdd_reg_lm()` applies the following TSLS procedure: treatment is predicted using $W_i$ and the cutoff dummy $Z_i$, the instrumental variable, in the first stage regression. The fitted values from the first stage regression are used to obtain a consistent estimate of the treatment effect using the second stage where the outcome $Y$ is regressed on the fitted values and the running variable $W$.

```{r}
# estimate the Fuzzy RDD
data <- rdd_data(d$Y, d$W, 
                 cutpoint = 0, 
                 z = d$treatProb)

frdd_mod <- rdd_reg_lm(rdd_object = data, 
                       slope = "same")
frdd_mod %>% render_summary()

```

The estimate is close to $2$, the population treatment effect. We may call `plot()` on the model object to obtain a figure consisting of binned data and the estimated regression function. 

```{r}
# plot estimated FRDD function
plot(frdd_mod, 
     cex = 0.5, 
     lwd = 0.4,
     xlim = c(-4, 4),
     ylim = c(-3.5, 5),
     xlab = "W",
     ylab = "Y")
```

What if we used a SRDD instead, thereby ignoring the fact that treatment is not perfectly determined by the cutoff in  $W$? We may get an impression of the consequences by estimating an SRDD using the previously simulated data.


```{r}
# estimate SRDD
data <- rdd_data(d$Y, 
                 d$W, 
                 cutpoint = 0)

srdd_mod <- rdd_reg_lm(rdd_object = data, 
                       slope = "same")
srdd_mod %>% render_summary()

```

The estimate obtained using a SRDD is suggestive of a substantial downward bias. In fact, this procedure is inconsistent for the true causal effect so increasing the sample would not alleviate the bias.
